
\documentclass[12pt]{article}
\usepackage[margin=1in]{geometry}

\input{../../../../../variables.tex}


\title{%
    \Huge Abstract Algebra \\
    \large by \\
    \Large Dummit and Foote \\~\\
    \huge Part 1: Group Theory \\
    \LARGE Chapter 3: Quotient Groups and Homomorphisms \\
    \Large Section 1: Definitions and Examples
}
\date{2024-04-12}
\author{Michael Saba}

\begin{document}
\pagenumbering{gobble}
\maketitle
\newpage
\setlength{\parindent}{0pt}
\pagenumbering{arabic}

This section introduces the concept of a quotient group,
which, like a subrgoup, allows us to obtain a smaller group
from another group. 
Like subgroups, we can use quotient groups of a group to study
its structure.
We can often combine information about subgroups and quotient
groups in order to derive theorems about groups. \\

\subsection*{Quotient Groups}

The study of the quotient groups of some group $G$
is essentially the study of the homomorphisms of $G$. \\
Suppose that we have a homomorphism $\varphi$
from $G$ onto another group $H$.
We know that the fibers of any mapping $\varphi$
are the sets of elements in $G$ that map to the same element
in $H$.
We will call $X_a$ the set of elememts that map to $a$
where $a \in H$,
and we say that it is the fiber lying over $a$. \\
Because the elements being mapped to by $\varphi$
are in a group $H$,
the group operation $H$ provides a way of multiplying them.
This suggests that we can perhaps define an operation
on the fibers themselves (the sets) that follows
the group operation of $H$;
that means that if $a, b \in H$ such that $ab = c$,
then we could define $X_a \cdot X_b$ to be $X_{ab} = X_{c}$.
This allows us to create a group out of the fibers of $G$. \\
We first need to prove that this is indeed a group:
\begin{itemize}[label=$\diamond$]
    \item 
        The operation is associative because
        the operation is associative in $H$.
    \item
        The identity is the fiber over the identity of $H$,
        $X_1$.
        This is because $X_1 \cdot X_a = X_{1 \cdot a} = X_{a}$
        and $X_a \cdot X_1 = X_{a \cdot 1} = X_{a}$.
    \item
        The inverse of a fiber $X_a$ is the fiber $X_{a^{-1}}$,
        since $X_a \cdot X_{a^{-1}} = X_{aa^{-1}} = X_1$.
    \item
        We know that for any two fibers $X_a$ and $X_b$,
        $X_a \cdot X_b = X_{ab}$.
        In order for the group to be closed under this operation,
        we would need to guarantee that $ab$ has a fiber
        over $\varphi$.
        This is true because, as we know,
        the image of a group $G$ over a homomorphism
        $\varphi: G \to H$
        is closed under the operation of $H$.
        This is because the image of $G$ is itself a group
        (since it is a subgrouo of $H$, 
        as shown in exercise 1.1.6.13,
        and as we will show later in this section).
        Thus the group is closed under its operation.
\end{itemize}
The group defined by the fibers of some homomorphism $\varphi$
from $G$ to $H$
is known as a \textbf{quotient group} of $G$. \\
Since the mutliplication of the fibers is defined
from the multiplication of the elements in $H$,
it is clear to see that the quotient group is isomorphic
to the image of $G$ under the homomorphism $\varphi$
(the elements of $H$ which have non-empty fibers in $G$). \\

We can think of this group as a way to abstract away
some class of elements from a group, and focus on the structure
as it relates to some other class of elements. \\
For example, there exists a homomorphism between any
Dihedral group $D_{2n}$, and the group of reflections $\zbb_2$. \\
The fibers of the $D_{2n}$ in this case would be the
two sets of elements that represent all rotations that
are either reflected or not. We've abstracted away rotations,
and instead gotten a group with just the reflections. \\

Now that we have some intuition of what quotient groups are,
we will show an alternative way of defining them that motivates
why we call them quotient groups. \\

Suppose we have a homomorphism $\varphi: G \to H$.
The fiber $X_1$ lying over the identity is called
the kernel of $\varphi$.
Basically, it is the set
\[ \{ g \in G \mid \varphi(g) = 1 \} \]
We can denote it using $\ker \varphi$. \\
This is related to the idea of the kernel of a group action.
We can recall that if a group $G$ is acting
on a a subset $S$ of $G$,
then the kernel of the action is the set of elements
in $G$ that satbilize all elements in $S$:
\[ \{ g \in G \mid g \cdot s = g \text{ for all } s \in S \}  \]
We can also recall that all group actions
from a group $G$ onto a set $A$
are themselves just a homomorphism $\phi: G \to S_A$
called the permuation representation of the group action
(since each group element essentially permutes $A$).
The kernel of the action of $G$ on its subset $S$
is the set of elements that don't permute $S$,
and send each of its element to themsleves.
So if we consider the permutation representation of this action
$\phi: G \to S_S$,
then the kernel of the action is the set of elements of $G$
that $\phi$ maps to the identity permutation
(the one that doesn't permute $S$),
which is exactly how we defined the kernel of a homomorphism
such as $\phi$. \\
Thus the kernel of group action and the kernel of homomorphism
are eseentially the same idea. \\

We will now prove some properties of homomorphisms
and their fibers.
Suppose that $\varphi: G \to H$ is a homomorphism:
\begin{itemize}[label=$\diamond$]
\item 
    $\varphi(1_G) = \varphi(1_H)$.
    This is true because
    $\varphi(1_G) = \varphi(1_G1_G) = \varphi(1_G)\varphi(1_G)$.
    So by the law of cancellation,
    it must be that $\varphi(1_G) = 1_H$
    (we mutltiply both sides by $\varphi(1_G)^{-1}$).
\item     
    $\varphi(g^{-1}) = \varphi(g)^{-1}$
    for any $g \in G$.
    This is true because
    $\varphi(1_G) = \varphi(gg^{-1}) 
    = \varphi(g)\varphi(g^{-1}) = 1_H$
    (by the first proposition we proved).
    Multiplying both sides by $\varphi(g)^{-1}$,
    we get $\varphi(g^{-1}) = \varphi(g)^{-1}$.
\item
    $\varphi(g^n) = \varphi(g)^n$ for all $n \in \nbb$.
    We can prove this using induction: \\
    \textbf{Basis Step:} We know that 
    $\varphi(g^1) = \varphi(g)^1$. \\
    \textbf{Inductive Hypothesis:} Assume that
    $\varphi(g^n) = \varphi(g)^n$. \\
    \textbf{Inductive Step:} We have:
    $ \varphi(g^{n+1}) = \varphi(gg^n)
    =  \varphi{g}\varphi(g^n) =  \varphi(g)^{n+1}$.
    We can then use the proof of the second part
    to extend this to negative exponents as well.
\item 
    $\ker \varphi$ is a subgroup of $G$. This is true
    because $1_G \in \ker \varphi$ by definition,
    the group operation is the same (associativity),
    and the following equation holds for any 
    $x, y \in \ker \varphi$:
    \[ \varphi(xy\inv) = \varphi(x)\varphi(y\inv)
    = \varphi(x)\varphi(y)\inv = 1_H1_H\inv = 1_H \]
    Which means that $xy\inv \in \ker \varphi$.
\item 
    $\im(\varphi)$, the image of $G$ under $\varphi$,
    is a subgroup of $H$. \\
    This is true for all homomorphisms and their co-domains.
    We already know that $1_H$ is part of the image since
    it is the image of $1_G$. The co-domain has the
    same group operator as $H$ itself, so it is associative.
    And finally, if $x, y \in \im(\varphi)$,
    then $x = \varphi(a)$ and $y = \varphi(b)$,
    where $a, b \in G$ are not necessarily distinct.
    And $\varphi(ab\inv) = \varphi(a)\varphi(b)\inv = xy\inv$,
    so $xy\inv$ is in the codomain.
\end{itemize}
With that we can define some terminology. \\

Now, if we have a homomorphism:
\[ \varphi: G \to H \]
Then we will define the quotient group of fibers
to be $G/K$, where where $K$ is the kernel of the homomorphism. \\
This is because we can think of the fibers as being
the quotients we get from dividing the group by the kernel.
For example, in the $D_{2n}$ example cited above, we get the
group of reflections by dividing out by the group of rotations,
which in this case is the kernel. \\

Notice that the other fibers will share the same structure
as the kernel. We say they are translates of the kernel. \\
For example, in the $D_{2n}$ case, the other fiber is the
set of reflected rotations. \\
However, out of all the fibers, only the kernel contains the
identity, and is therefore the only fiber that is also a group. \\

Because of that, and because it is the only one made up
purely of the operation we are trying to abstract (in the
$D_{2n}$ case, the rotation), it's the one we use when dividing
$G$. It wouldn't make sense to divide by the
set of reflected rotations,
since those contain reflections (and aren't a group). \\

\subsection*{Cosets}

Now that we have defined what a quotient group is,
and gotten an intuition as to what it is and what it contains,
we can move on to defining it without using homomorphisms. \\
This is useful, since the group of fibers itself is just
a partition of the elements of a group $G$, such that the sets
form a group consistent with the group operation of $G$
and the way we partitioned the elements.
The homomorphism itself is not a necessary part of that
definition. \\

Recall that we needed the homomorphism from $G$ to $H$,
as the group operation of $H$ told us how to combine the
fibers and turn them into a group. \\
Instead, we will now use the group operation of $G$
in order to define the operation. \\
So if we have two fibers $X_a$ and $X_b$, such that
$g \in X_a$ and $h \in X_b$ are two elements, one in each 
fiber, we will define the group operation to be:
\[ X_a \cdot X_b = C \]
Where $C$ is the group that contains $gh$. \\

This operation has to be well defined, that is, indepdent
of the choice of representatives $g$ and $h$. \\
The concept of well-defintion is usually used when functions
map sets to other sets using representatives from the sets,
and in this case, the function is the group operation. \\

If we know that $X_a$ and $X_b$ are fibers of a homomorphism,
this is trivial. \\
We know that $\varphi(gh) = \varphi(g)\varphi(h) = X_aX_b$
for any choice of representatives. \\

\end{document}
