

\documentclass[12pt]{article}

\usepackage[margin=1in]{geometry}

\input{../../variables.tex}

\title{%
    \Huge Numerical Algorithms \\
    \Large Lecture X
}
\date{2025-05-21}
\author{Michael Saba}

\begin{document}
\pagenumbering{gobble}
\maketitle
\newpage
\setlength{\parindent}{0pt}
\pagenumbering{arabic}

\subsection*{Code}

In this lecture, we look at the code
for a ray tracer. \\

First we have gamma correction of the pixels,
which is done at the end, and is only done
for visual purposes. \\

In Monte-Carlo integration, instead
of shooting several rays each
frame, we calculate the integral
using several samples and then average out. \\
This is far more efficient, and is done
for each term $T^kL_e$. \\

However, in practice,
instead of averaging out at each
bounce $T^kL_e$, we just choose a single sample,
and follow a single ray's path.
We then average out all of the paths
later, when we have several paths. \\

If we have an average:
\[ S_{n-1} = \sum_{i = 1}^{n-1} r_{i} \]
Then:
\[ S_{n} = \sum_{i = 1}^{n} r_{i} \]
But we don't need to store or recalulate
each new average, because:
\[ S_{n} = (n-1)S_{n-1} + r_n \]
We don't generate and average out several
ray paths in a single frame; we do it
frame by frame, and use the above method to
average out. \\
When we move the camera, we have to reset
and start from the beginning. \\

We then build the scene using ray intersections. \\

For Monet-Carlo, we need a random number generator;
but we do Monte-Carlo for each pixel,
and we need a different random number generator
for each, so we have to use a seed,
which is a hash from the pixel coordinate. \\

We use one sample per pixel in Monte-Carlo. \\

We shoot a ray through several points in our
pixel, and see what it intersects. \\

At this point, we can do several color options. \\

The first is just black and white, 
if there is or isn't intersection. \\

Another is depth, which gets brighter the closer
the intersection is. \\

Another is normal coloring. \\
Don't forget to rescale normal, since it's
-1 to 1 (should be 0 to 1). \\

Now for the ray tracer, we have a function
that send a ray and returns the point it hit. \\
If we hit nothing, return background color,
else if we hit something, we generate the normal,
and we generate random rays in the upper hemisphere
around the normal. We then set the new origin
and direction of the ray, and send it again. \\
In practice we only send one ray (have several
series consiting of one ray, which we all
average out). \\
Also when we intersect, we add an epsilon
along the normal, to ensure our intersected
point isn't inside the geometry. \\

Before we send the new ray, we
add the $L_e$ of the point we intersect
if it exists, and apply Monte-Carlo integration,
using a double integral with variable change
(unit hypercube range). \\
So we have:
\[ L_e + L_e \times \text{BRDF} \]
Note that the determinants of the Jacobians,
and the $\cos(\theta)$ terms cancel out;
this is by design of our map. \\
This gets rid of variance/noise in the function,
which is useful. 
We call it a cosine weighted hemisphere
for this reason; it cancels the terms. \\
Notice that we don't have an integral
since we only want one sample per ray
intersection as explained. \\
We then continue, adding a term $T^kL_e$
for each new bounce,
and add it to the accumulated color
($L_e + TL_e + T^2L_e, \dots$).
For $T^k$, we would need to multiply
$L_e$ by an accumulated product
(called throughput). \\
If we eventually hit nothing,
it just becomes the background color
multiplied by the accumulated color up
to this point (or background texture).
If it hits the limit of bounces (since bounces
contribute less and less due to the BRDF
absorbing light). \\
If some surface we hit has no emitted light,
it adds nothing to the accumulated light. \\

Small note about Monte-Carlo;
in theory, when we integrate over several
rays, the throughput, which is BRDF
and other terms such as the $\cos$,
would decrease with more bounces, to conserve
energy. \\
However, in Monte-Carlo integration,
instead of sending several rays of light
each bounce, we have one ray of light
that we follow for a whole path,
and then average out several paths. 
However, each individual path may not converge,
and the throughput may continue increase;
the average has to converge, but any one
individual path may increase in energy. \\

Note that we can do a coloring
with just a single bounce, where we don't
have to save the accumulated color
and the accumulated product. \\

Note that when we do the Hemisphere map
to spherical coordinates, we are calculating
the coordinates of a Hemisphere pointing up,
but the geometry may have a hemisphere in any direction;
the answer is that because $R^TR = I$,
rotations being orthogonal,
the Jacobian is calculated the same way;
translations and rotations don't affect the
Jacobian. \\

The mapping that turns $[0, 2\pi] 
\times [0, \sfrac{\pi}{2}]$
does not have to be linear, where
$\phi$ and $\theta$
each only depend on $u$ and $v$,
and where we only apply a linear map. \\
But it does make things simpler. \\
Also note that for this transformation,
the Jacobian is square, do we can just calculate
$|\det(J)|$. \\

If we are asked to find a Monte-Carlo estimator of
an intergral:
\[ \int_\Omega f(x) dx \]
We need to change the variables,
and compute the integral as an average:
\[ \dfrac{1}{n}\sum_{i=1}^n f(T_1(T_2(u, v)))
\cdot |\det(T_1)| \cdot |\det(T_2)| \]
Where we removed the outer integrals
from $0$ to $1$ as per Monte-Carlo. \\
Note that in practice, we don't average
out at each bounce, but follow one ray
and then average out all the ray paths,
though we do it frame by frame. \\

\newpage

\end{document}