


\documentclass[12pt]{article}

\usepackage[margin=1in]{geometry}

\input{../../variables.tex}

\title{%
    \Huge Numerical Algorithms \\
    \Large Lecture XIII
}
\date{2025-06-02}
\author{Michael Saba}

\begin{document}
\pagenumbering{gobble}
\maketitle
\newpage
\setlength{\parindent}{0pt}
\pagenumbering{arabic}

\subsection*{Inverse-Discrete Sampling}

Inverse sampling was done in the continuous case,
where we wanted to get rid of a term $p(x)$
in the intgeral. \\

In the discrete case, we have the probabilities:
\[ \{p_1, p_2 \dots p_n \} \]
Which sum up to $1$. \\
This is the discrete analog to $p(x)$. \\

If the sum of $p_i$ is $M \neq 1$, we just divide
all probabilities by $M$. \\

We define:
\[ P_i = \sum_{k=1}^i p_k \]
And we map the intervals:
\[ \brac{ P_i,  P_{i+1}} \to i+1 \]
to these indices. \\
This is the integration in the continuous case. \\

We then try to fit $u \in [0, 1]$ 
into one of the buckets we have,
using binary search for efficiency
(we can also use linear search), which is our
inversion that we did in the continuous case. \\

The funciton is not injective (piece-wise constant function), 
which is why we can't
invert it, which is why we have buckets that map to
indexes. \\

For the $2$ dimensional case, we have rows and columns,
we can then sum all the rows (or columns),
and get a $1$ dimensional array. \\

Video: My favorite samples | SIGGRAPH courses 1:42:20. \\

\newpage

\subsection*{Heat diffusion}

Generally particles move from higher to lower concentration
parts; as time goes on, the enture image blurs, as
the heat diffuses in space, and the heat distributes. \\
The color at the steady state
(equilibrium) is a constant, mean, color. \\

Works not just for heat, but for pressure, and for
mass... \\

We can express the heat
in some region $\Om$ at time $t$
as $Q(t, \Om)$, where:
\[ Q(t, \Om) = \int_\Om p(t, x, y, z) 
\; dx \; dy \; dz
= \int_\Om p(t, x, y, z) \; dV \]
Where $p$ is the density at a point. \\

The change in heat in the region $\Om$
with respect to the time is given by:
\[ \partialdd{}{t} \int_\Om p(t, x, y, z) \; dV \]
It can be decomposed into the generated
heat in the region minus the heat
leaving the region $\Om$,
which means the heat leaving in the direction
of the normal at each point
on the boundary of the region $\partial \Om$:
\[ \partialdd{}{t} \int_\Om p(t, x, y, z) \; dV
= \int_\Om F \; dV
- \int_{\partial \Om} j \cdot n \; dA \]
Where the heat leaving the region is
given by a flux integral (dot product
of vector field with normal). \\

Here we have $j$ as the vector
field of change of density $p$:
\[ j = D\nabla p \]
Here, the matrix $D$
decides how much change we have in each
direction. \\
However, by Fick's law,
we can assume we have constant change
in each direction, which means that
$D = kI$, so we can write:
\[  \]


\end{document}
